---
title: "Untitled"
author: "William Hemstrom"
date: "1/18/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(snpR); library(ggplot2); library(gridExtra); library(data.table); library(openxlsx); library(ggtern)

nsites <- 60589
wb <- openxlsx::createWorkbook(creator = "William Hemstrom")
openxlsx::addWorksheet(wb, "Table S2")
openxlsx::addWorksheet(wb, "Table S3")
openxlsx::addWorksheet(wb, "Figure 3")
openxlsx::addWorksheet(wb, "Figure 4")
openxlsx::addWorksheet(wb, "Figure 5")
openxlsx::addWorksheet(wb, "PCA and NGSadmix")
openxlsx::addWorksheet(wb, "thetas")

```
# Read in and prep data
## Read in the genotype and metadata

First, read in and clean the basic metadata.
```{r, echo=FALSE, message=FALSE, results='hide'}
# meta
sample_meta <- read.table("metadata.txt", header = T)
sample_meta$stream <- paste0("str_", sample_meta$stream)
sample_meta$stream[sample_meta$watershed == "LHU" & sample_meta$stream == "str_INC"] <- "str_INK"
## merge pops ran together during pva
pva_merge_key <- data.frame(old = c("SEC", "TRB", "SHE", "SNO", "MAE", "MAR", "MRB", "MAU"), new = c(rep("GRP", 4), rep("MRG", 4)))
for(i in 1:nrow(pva_merge_key)){
  sample_meta$stream <- gsub(pva_merge_key[i,1], pva_merge_key[i,2], sample_meta$stream)
}
## fix year for some MRG samples
sample_meta[sample_meta$stream == "str_MRG",]$year <- 2016

```

Read in the hybridization info, then merge with meta. Need to use the key to adjust for name changes.

```{r, echo=FALSE, message=FALSE, results='hide'}
# read
hyb <- read.csv("YCT_RBT_hybrid_assessment.csv")
hyb_key <- openxlsx::read.xlsx("Table S1 20210111.xlsx", sheet = 2)
hyb <- merge(hyb, hyb_key, by.x = "sample_ID", by.y = "New.ID")

# merge
comb_meta <- merge(sample_meta, hyb[,c(15, 11:13)], by.x = "ID", by.y = "Final.ID")
```

Read in the genotypes, convert to snpRdata.

```{r, echo=FALSE, message=FALSE, results='hide'}
genos <- read.table("bamlist2_baits_geno.geno", stringsAsFactors = F)

snp_meta <- genos[,1:2]
colnames(snp_meta) <- c("chr", "position")
dat <- import.snpR.data(genos[,-c(1:2)], snp_meta, comb_meta)

```

Remove samples that have a %LCT less than 99%. Filter SNPs.

```{r, echo=FALSE, message=FALSE, results='hide'}
# filter bad individuals
dat <- dat[,-which(sample.meta(dat)$LCT < .99)]

# save filtering info
n <- table(paste0(sample.meta(dat)[,"stream"], ".", sample.meta(dat)[,"year"]))
n <- as.data.frame(n)
colnames(n) <- c("subfacet", "Final n")

# 
# # filter for HWE per pop
# dat <- filter_snps(dat, hwe = 0.000001, hwe_facets = "stream.year") # removing 5 loci


```
# Calculate and merge together statistics
## Calculate basic diversity statistics

Ne via NeEstimator, ho, he, het_hom, and Tajima's D. Uncomment to run.

```{r, echo=FALSE, message=FALSE, results='hide'}
# dat <- calc_ne(dat, "stream.year", chr = "chr", NeEstimator_path = "C://usr/bin/Ne2-1.exe", methods = c("LD"))
# dat <- calc_ho(dat, c("stream.year"))
# dat <- calc_pi(dat, "stream.year")
# dat <- calc_het_hom_ratio(dat, c("stream.year"))
# dat <- calc_tajimas_d(dat, "stream.year.chr", par = 6)
# dat <- calc_fis(dat, "stream.year")
# saveRDS(dat, "processed_snpRdata.RDS")
dat <- readRDS("processed_snpRdata.RDS")
```

## Run Colony
If this needs to be run still, uncomment. Otherwise, just read in the results.

```{r, results='hide'}
# dir.create("colony")
# setwd("colony/")
# 
# pops <- unique(sample.meta(dat)[,3:4])
# 
# #=========run and source sibs==============
# out <- vector("list", nrow(pops))
# 
# for(i in 1:nrow(pops)){
#   tdat <- dat[stream.year = paste0(pops[i, 1], ".", pops[i, 2])]
#   if(nsamps(tdat) <= 1 | nsnps(tdat) <= 10){
#     out[[i]] <- data.frame(offspringID1 = character(), ofspringID2 = character(), Probability = numeric())
#   }
#   else{
#     tdat <- filter_snps(tdat, maf = 0.01)
#     if(nsamps(tdat) <= 1 | nsnps(tdat) <= 10){
#       out[[i]] <- data.frame(OffspringID1 = character(), OffspringID2 = character(), Probability = numeric())
#     }
#     else{
#       dir.create(paste0(pops[i,], collapse = "_"))
#       setwd(paste0(pops[i,], collapse = "_"))
#       write_colony_input(tdat, sampleIDs = "ID")
#       call_colony("colony/colony_input.dat", colony_path = "C:/usr/bin/Colony/colony2s.exe", update_bib = FALSE)
#       out[[i]] <- read.csv("colony/colony_input.FullSibDyad", header = T, stringsAsFactors = F)
#       setwd("../")
#     }
#   }
# }
# empties <- which(lapply(out, nrow) == 0)
# cout <- dplyr::bind_rows(out[-empties])
# 
# write.table(cout, "colony_sibs.txt", quote = F, sep = "\t", row.names = F)
# setwd("../")

#===============harvest and save ne===================
nef <- list.files(path ="colony/", pattern = "Ne", recursive = T)
out <- data.frame(pop = gsub("/.+", "", nef),
                  ne.rand = numeric(length(nef)),
                  lcl.rand = numeric(length(nef)),
                  ucl.rand = numeric(length(nef)),
                  ne.nonrand = numeric(length(nef)),
                  lcl.nonrand = numeric(length(nef)),
                  ucl.nonrand = numeric(length(nef)))

for(i in 1:length(nef)){
  col_ne <- readLines(paste0("colony/", nef[i]))
  col_ne <- gsub(" ", "", col_ne)
  col_ne <- gsub("^.+=", "", col_ne)
  out[i,-1] <- as.numeric(col_ne[c(5,6,7,12,13,14)])
}

out[out == 2147483647] <- Inf
# write.table(out, "colony_ne.txt", quote = F, sep = "\t", row.names = F)
col_ne <- out
```


## Fetch and merge statistics

Tajima's D first:
```{r, results='hide'}
tajimas_D <- get.snpR.stats(dat, "stream.year.chr", "tajimas_d")$weighted.means
tajimas_D <- dplyr::filter(tajimas_D, snp.subfacet == ".OVERALL_MEAN")
tajimas_D <- tajimas_D[,-c(1, 3, 4, 7)]
colnames(tajimas_D)[2:3] <- gsub("weighted_mean_", "", colnames(tajimas_D)[2:3])
tajimas_D$ws.theta <- tajimas_D$ws.theta/nsites # devide by total number of sequenced sites to change the denominator
tajimas_D$ts.theta <- tajimas_D$ts.theta/nsites # devide by total number of sequenced sites to change the denominator
```

Other basic stats:

```{r, results='hide'}
ne <- get.snpR.stats(dat, "stream.year", "ne")$pop
colnames(ne)[2] <- "subfacet"
diversity <- get.snpR.stats(dat, "stream.year", c("pi", "ho", "het_hom_ratio", "fis"))$weighted.mean
```

# PCA and NGSadmix plots

See Hyb_info.Rmd for code

```{r}
# read in
cluster_plots <- readRDS("cluster_plots.RDS")

l1 <- ggpubr::get_legend(cluster_plots$pca)
l2 <- ggpubr::get_legend(cluster_plots$ngs)

# get the total number of hybrids per pop...
check <- cluster_plots$pca$data
check <- check[-which(check$locale %in% c("YCT_YCT", "RBT_RBT")),]
check$stream <- gsub("str_BAT", "str_NFB", check$stream)
for(i in 1:nrow(pva_merge_key)){
  check$stream <- gsub(pva_merge_key[i,1], pva_merge_key[i,2], check$stream)
}
check$subfacet <- paste0(check$stream, ".", check$year)
check$subfacet <- gsub("str_MRG.2017", "str_MRG.2016", check$subfacet)
check <- as.data.table(check)


check[,bad := LCT < 0.99]
n_rem <- table(check$subfacet[which(check$bad)])
n_rem <- as.data.frame(n_rem)
colnames(n_rem) <- c("locale", "Number of Hybrids (%LCT < .99)")

merge_and_add <- function(x, tab, new.name){
  tab <- as.data.frame(tab)
  tab$locale <- rownames(tab)
  colnames(tab)[1] <- c(new.name)
  return(merge(tab, x, "locale", all = TRUE))
}

n_rem <- merge_and_add(n_rem, tapply(check$LCT, check$subfacet, mean), "%LCT before filtering")
n_rem <- merge_and_add(n_rem, tapply(check$YCT, check$subfacet, mean), "%YCT before filtering")
n_rem <- merge_and_add(n_rem, tapply(check$RBT, check$subfacet, mean), "%RBT before filtering")

n_rem <- merge_and_add(n_rem, with(check[-which(check$bad),], tapply(LCT, subfacet, mean)), "%LCT after filtering")
n_rem <- merge_and_add(n_rem, with(check[-which(check$bad),], tapply(YCT, subfacet, mean)), "%YCT after filtering")
n_rem <- merge_and_add(n_rem, with(check[-which(check$bad),], tapply(RBT, subfacet, mean)), "%RBT after filtering")

n_rem <- merge(n_rem, unique(check[, c("watershed", "stream", "year", "subfacet")]), by.x = "locale", by.y = "subfacet")


# note the numbers removed/percent hybrid before and after.

gridExtra::grid.arrange(cluster_plots$pca + ggplot2::guides(color = "none"), l1, 
             cluster_plots$ngs + ggplot2::guides(fill = "none", color = "none"), l2,
             nrow = 2, ncol = 2,
             layout_matrix = cbind(c(1, 3), c(2, 4)),
             widths = c(1, .2))
openxlsx::insertPlot(wb, "PCA and NGSadmix", width = 11, height = 8.5)

```

# merge all of the stats together

```{r, results='hide'}
# merge Tajima's D and ne with diversity stats
stats <- merge(diversity, ne, by = c("facet", "subfacet"))
stats <- merge(stats, tajimas_D, by = "subfacet")

# sample sizes
stats <- merge(stats, n, by = "subfacet", all = TRUE)

# hyb percents
stats <- merge(stats, n_rem[,1:8], by.x = "subfacet", by.y = "locale", all = TRUE)
stats$`Final n`[which(is.na(stats$`Final n`))] <- 0 

# combine and reformat
stats$weighted_mean_pi <- stats$weighted_mean_pi/nsites
stats$weighted_mean_ho <- stats$weighted_mean_ho/nsites
colnames(stats)[which(colnames(stats) == "weighted_mean_pi")] <- "pi"
colnames(stats)[which(colnames(stats) == "weighted_mean_ho")] <- "ho"
colnames(stats)[which(colnames(stats) == "weighted_mean_fis")] <- "fis"

# add scaled tsd
stats$theta_diff <- (stats$ts.theta - stats$ws.theta)/rowMeans(stats[,c("ts.theta", "ws.theta")])




```

Merge in colony info
```{r, results='hide'}
# merge in colony Ne info
col_ne$pop <- gsub("_", ".", col_ne$pop)
col_ne$pop <- gsub("str.", "str_", col_ne$pop)
stats <- merge(stats, col_ne, by.x = "subfacet", by.y = "pop", all.x = TRUE)
```

Merge in PVA data

```{r, results='hide', warning=FALSE}
#===========newer PVA data=============
new_PVA <- readr::read_csv("Tables_Figs_Ave N_ALL for Will 1-4-22.csv")
new_PVA <- as.data.table(new_PVA)
mod.cols <- 1:ncol(new_PVA)
new_PVA[,(mod.cols) := lapply(.SD, function(x) gsub("#N/A", NA, x)), .SDcols = mod.cols] # fix NA labels
need.num.cols <- c(3:9,11:ncol(new_PVA))
new_PVA[,(need.num.cols) := lapply(.SD, as.numeric), .SDcols = need.num.cols]
old_PVA <- readr::read_csv("Table_Thetas_hyb_PVA_03062020.csv")
colnames(old_PVA)[which(colnames(old_PVA) == "Genetic Year")] <- "year"
old_PVA <- as.data.table(old_PVA)
new_PVA <- merge(new_PVA, old_PVA[,c(1:8)], by = c("Basin", "Creek", "year"), all = TRUE)
## merge pops ran together during pva
for(i in 1:nrow(pva_merge_key)){
  new_PVA$Stream_code <- gsub(pva_merge_key[i,1], pva_merge_key[i,2], new_PVA$Stream_code)
}
was_merged <- which(new_PVA$Stream_code == "str_MRG")
new_PVA[was_merged]$Creek <- "Merged Marys River"
new_PVA <- new_PVA[-was_merged[2:3],]
new_PVA[, subfacet := paste0(Stream_code, ".", year)]
new_PVA <- as.data.frame(new_PVA)

stats <- merge(stats, new_PVA, by = "subfacet", all = TRUE)


```
## Output table

```{r}
Table.S2 <- stats[, -c(which(colnames(stats) %in% c("snp.facet", "snp.subfacet", "weighted_mean_Het/Hom", "LDNe_0.02", "Group",
                                                    "Sample_ID_Prefix", "popname", "popnum", "year_genetics")),
                       grep("CIp", colnames(stats)),
                       grep("02$", colnames(stats)))]


openxlsx::writeData(wb, "Table S2", x = Table.S2, keepNA = T)



```


# Statistical analysis vs. PVA data

Remove the bad PVA pops.

```{r}

bad.locales <- c("str_CMC.2016", "str_AND.2016", "str_LIN.2017", "str_LEE.2016", "str_NFG.2009", "str_PEC.2018", "str_MEI.2017", "str_POL.2016", "str_BYD.2015")
stats <- stats[-which(stats$subfacet %in% bad.locales),]

```


## Figure 3
```{r}
# basic correlations:
## melt data
# year.matches <- which(abs(stats$year  - stats$PVA.Abund.Year) <= 5 & stats$`Final n` >= 5)
year.matches <- which(!is.na(stats$PVA.Abund.Year) & stats$`Final n` >= 5)
cor_dat <- stats[year.matches,]
cor_dat <- reshape2::melt(cor_dat[,c("subfacet", "Basin", "Creek", "year", "pi", "ho",
                                     "ts.theta", "ws.theta", "theta_diff", "PVA.Extinction", "PVA.Abundance.50", "N_harmean")], 
                          id.vars = c("subfacet", "Basin", "Creek", "year", "PVA.Extinction", "PVA.Abundance.50", "N_harmean"))

## change variable names to be pretty
pretty_names_tab <- c(ho = "H[o]",
                      pi = "pi",
                      ws.theta = "theta[W]",
                      ts.theta = "theta[T]",
                      theta_diff = "theta[diff]",
                      fis = "F[IS]",
                      ne.rand = "Ne[Colony]")


vlabeller <- as_labeller(pretty_names_tab, default = label_parsed)

## plots
cor_plot_ab <- ggplot(cor_dat, aes(y = log10(PVA.Abundance.50), x = value)) + 
  facet_wrap(~variable, scales = "free_x", labeller = vlabeller, strip.position = "bottom") + 
  geom_point() + theme_bw() + 
  theme(strip.background = element_blank(), axis.title.x = element_blank(), 
        strip.placement = "outside", axis.text.x = element_text(size = 10, angle = 90), 
        strip.text = element_text(size = 12),
        axis.text.y = element_text(size = 10),
        panel.spacing.x = unit(1, "lines")) + ggtitle("A") + ylab("log10(Estimated PVA Abundance)")



cor_plot_ext <- ggplot(cor_dat, aes(y = log10(PVA.Extinction), x = value)) + 
  facet_wrap(~variable, scales = "free_x", labeller = vlabeller, strip.position = "bottom") + geom_point() +
  theme_bw() + 
  theme(strip.background = element_blank(), axis.title.x = element_blank(), 
        strip.placement = "outside", axis.text.x = element_text(size = 10, angle = 90), 
        strip.text = element_text(size = 12),
        axis.text.y = element_text(size = 10),
        panel.spacing.x = unit(1, "lines")) + ggtitle("B") + ylab("log10(% Estimated PVA Extinction Risk)")

cor_plot_harmean <- ggplot(cor_dat, aes(y = log10(N_harmean), x = value)) + 
  facet_wrap(~variable, scales = "free_x", labeller = vlabeller, strip.position = "bottom") + geom_point() +
  theme_bw() + 
  theme(strip.background = element_blank(), axis.title.x = element_blank(), 
        strip.placement = "outside", axis.text.x = element_text(size = 10, angle = 90), 
        strip.text = element_text(size = 12),
        axis.text.y = element_text(size = 10),
        panel.spacing.x = unit(1, "lines")) + ggtitle("B") + ylab("log10(Estimated Harmonic Mean Population Size)")


# rf

rfstats <- stats[year.matches,]

## clean
rfstats$Basin <- as.factor(rfstats$Basin)
rfstats$Creek <- as.factor(rfstats$Creek)
rfstats$PVA.Extinction <- scale(rfstats$PVA.Extinction)
rfstats$PVA.Abundance.50 <- scale(rfstats$PVA.Abundance.50)
rfstats$N_harmean <- scale(rfstats$N_harmean)
rfstats$pi <- scale(rfstats$pi)
rfstats$ho <- scale(rfstats$ho)
rfstats$ws.theta <- scale(rfstats$ws.theta)
rfstats$ts.theta <- scale(rfstats$ts.theta)




## run a random forest against extinction risk
rfstats_ext <- rfstats[,c("PVA.Extinction", "ho", "pi", "ws.theta", "ts.theta")]


## full model
rf_ext <- ranger::ranger(dependent.variable.name = "PVA.Extinction", 
                         data = rfstats_ext, 
                         num.trees = 100000, 
                         mtry = ncol(rfstats_ext) - 1,
                         num.threads = 4, 
                         verbose = T, importance = "permutation")

pred_rf <- data.frame(obs = rfstats_ext$PVA.Extinction,
                      pred = rf_ext$predictions, statistic = "Extinction")

rf_ext$r.squared

## run a random forest against PVA.Abundance
rfstats_ab <- rfstats[,c("PVA.Abundance.50", "ho", "pi", "ws.theta", "ts.theta")]
rfstats_ab <- na.omit(rfstats_ab)



## full model
rf_ab <- ranger::ranger(dependent.variable.name = "PVA.Abundance.50", 
                        data = rfstats_ab, 
                        num.trees = 100000, 
                        mtry = ncol(rfstats_ab) - 1,
                        num.threads = 4,
                        verbose = T, importance = "permutation")

rf_ab$r.squared

pred_rf <- rbind(pred_rf,
                 data.frame(obs = rfstats_ab$PVA.Abundance.50,
                            pred = rf_ab$predictions,
                            statistic = "Abundance"))


## run a random forest against N_harmean
rfstats_harmean <- rfstats[,c("N_harmean", "ho", "pi", "ws.theta", "ts.theta")]
rfstats_harmean <- na.omit(rfstats_harmean)



## full model
rf_harmean <- ranger::ranger(dependent.variable.name = "N_harmean", 
                        data = rfstats_harmean, 
                        num.trees = 100000, 
                        mtry = ncol(rfstats_harmean) - 1,
                        num.threads = 4,
                        verbose = T, importance = "permutation")

rf_harmean$r.squared

pred_rf <- rbind(pred_rf,
                 data.frame(obs = rfstats_harmean$N_harmean,
                            pred = rf_harmean$predictions,
                            statistic = "Harmonic Mean N"))


## plot rf data
pred_rf$new_statistic <- ifelse(pred_rf$statistic == "Abundance", "log10(Estimated PVA Abundance)", ifelse(pred_rf$statistic == "Extinction", "log10(% Estimated PVA Extinction Risk)", "log10(Estimated Harmonic Mean Population Size)"))
pred_rf$new_statistic <- factor(pred_rf$new_statistic, levels = c("log10(Estimated PVA Abundance)", "log10(% Estimated PVA Extinction Risk)", "log10(Estimated Harmonic Mean Population Size)"))


# rf_p <- ggplot(pred_rf, aes(x = obs, y = pred)) + geom_point() +
#   theme_bw() + facet_wrap(~new_statistic, scales = "free") + theme(strip.background = element_blank()) +
#   ylab("Predicted") + xlab("Observed") + ggtitle("C")
rf_p <- ggplot(pred_rf[-which(pred_rf$statistic == "Abundance"),], aes(x = obs, y = pred)) + geom_point() +
  theme_bw() + facet_wrap(~new_statistic, scales = "free") + theme(strip.background = element_blank()) +
  ylab("Predicted") + xlab("Observed") + ggtitle("C")


# combine plots
# grid.arrange(cor_plot_ab + theme(plot.margin = unit(c(.4, .4, .4, .4), "cm")), 
#              cor_plot_ext + theme(plot.margin = unit(c(.4, .4, .4, .4), "cm")),
#              cor_plot_harmean + theme(plot.margin = unit(c(.4, .4, .4, .4), "cm")),
#              rf_p + theme(plot.margin = unit(c(.4, .4, .4, .4), "cm")),
#              layout_matrix = matrix(c(1,2,3, 4, 4, 4), byrow = T, nrow = 2),
#              heights = c(1, .5))
grid.arrange(cor_plot_ext + theme(plot.margin = unit(c(.4, .4, .4, .4), "cm")),
             cor_plot_harmean + theme(plot.margin = unit(c(.4, .4, .4, .4), "cm")),
             rf_p + theme(plot.margin = unit(c(.4, .4, .4, .4), "cm")),
             layout_matrix = matrix(c(1,2,3,3), byrow = T, nrow = 2),
             heights = c(1, .5))
openxlsx::insertPlot(wb, "Figure 3", width = 11, height = 8.5)


```

## Table S3, part 1

```{r}
#===============Table S3, part 1================
mod_stats <- stats[year.matches,]
mod_stats <- mod_stats[,c("Basin", "Creek", "year", "pi", "ho",
                          "ts.theta", "ws.theta", "theta_diff", "PVA.Extinction", "N_harmean")]
run_stats <- c("pi", "ho", "ts.theta", "ws.theta", "theta_diff")

# clean
mod_stats$Basin <- as.factor(mod_stats$Basin)
mod_stats$Creek <- as.factor(mod_stats$Creek)
mod_stats$N_harmean <- scale(mod_stats$N_harmean)
mod_stats$PVA.Extinction <- scale(mod_stats$PVA.Extinction)

models <- data.frame(stat = numeric(length(run_stats)),
                     Ext = numeric(length(run_stats)),
                     HM = numeric(length(run_stats)),
                     Ext.model = character(length(run_stats)),
                     n = numeric(length(run_stats)),
                     stringsAsFactors = F)

for(i in 1:length(run_stats)){
  # prep
  tdat <- mod_stats[,c("Creek", "Basin", "N_harmean", "PVA.Extinction", run_stats[i])]
  colnames(tdat)[5] <- "stat"
  bads <- which(is.na(tdat$stat) | is.infinite(tdat$stat))
  if(length(bads) > 0){
    tdat <- tdat[-bads,]
  }
  tdat$stat <- scale(tdat$stat)
  
  # run models, checking if the lmer worked for ext and running a basic lm if not
  cat("\n", i, "\n")
  cat("HM:\n")
  HM <- lmerTest::lmer(N_harmean ~ stat + (1|Basin) + (1|Basin:Creek), tdat)
  cat("Ext:\n")
  Ext <- try(lmerTest::lmer(PVA.Extinction ~ stat + (1|Basin) + (1|Basin:Creek), tdat), silent = T)
  out <- numeric(2)
  out[2] <- summary(HM)$coefficients[2,5]
  
  ## check if the lmer for Ext is good:
  goodlme <- T
  if(class(Ext) == "try-error"){goodlme <- F}
  else if(!is.null(summary(Ext)$optinfo$conv$lme4$code) | !is.null(summary(HM))){goodlme <- F}
  
  
  if(!goodlme){
    cat("Statistic: ", run_stats[i], " failed to converge with lmer.\n")
    Ext <- lm(PVA.Extinction ~ stat, tdat)
    out[1] <- summary(Ext)$coefficients[2,4]
    models[i, 4] <- "lm"
  }
  else if(goodlme){
    cat("Statistic: ", run_stats[i], " converged with lmer.\n")
    out[1] <- summary(Ext)$coefficients[2,5]
    models[i, 4] <- "glmm"
  }
  
  models[i,1] <- run_stats[i]
  models[i,2:3] <- out
  models$n <- nrow(tdat)
}

Table.S3 <- models
Table.S3$holm_Ext <- p.adjust(Table.S3$Ext, "holm")
Table.S3$holm_HM <- p.adjust(Table.S3$HM, "holm")

```

## Figure 4

```{r}
change_dat <- reshape2::melt(stats[,c("subfacet", "Basin", "Creek", "year", "pi", "ho",
                                            "ts.theta", "ws.theta", "theta_diff", "PVA.Extinction", "N_harmean")], 
                             id.vars = c("subfacet", "Basin", "Creek", "year", "PVA.Extinction", "N_harmean"))
change_dat <- change_dat[-which(is.na(change_dat$year)),]

change_dat$subfacet <- paste0(change_dat$Basin, "_", change_dat$Creek)
min_year <- tapply(change_dat$year, change_dat$subfacet, min, na.rm = T)
change_dat$min_year <- min_year[match(change_dat$subfacet, names(min_year))]
change_dat$year_delta <- change_dat$year - change_dat$min_year

change_plot <- ggplot(change_dat, aes(x = year_delta, y = value, color = subfacet)) +
  theme_bw() + facet_wrap(~variable, scales = "free_y", strip.position = "left", labeller = vlabeller) +
  geom_line(aes(group = subfacet), alpha = 0.5) +
  geom_point() + 
  theme(legend.position = "none", strip.background = element_blank(), 
        axis.title.y = element_blank(), strip.placement = "outside", 
        strip.text = element_text(size = 12), axis.text = element_text(size = 10)) +
  scale_color_viridis_d() + xlab("Years Since First Sample")
change_plot
openxlsx::insertPlot(wb, "Figure 4", width = 11, height = 8.5)
```

## Figure 5

```{r}
change_dat <- stats[year.matches,]
change_dat <- reshape2::melt(change_dat[,c("subfacet", "Basin", "Creek", "year", "pi", "ho", "ws.theta",
                                           "ts.theta", "theta_diff", "PVA.Extinction", "N_harmean")], 
                             id.vars = c("subfacet", "Basin", "Creek", "year", "PVA.Extinction", "N_harmean"))

change_dat$subfacet <- paste0(change_dat$Basin, "_", change_dat$Creek)
min_year <- tapply(change_dat$year, change_dat$subfacet, min, na.rm = T)
change_dat$min_year <- min_year[match(change_dat$subfacet, names(min_year))]
change_dat$year_delta <- change_dat$year - change_dat$min_year

stat_deltas <- data.frame(Creek = character(0), Basin = character(0), subfacet = character(0), variable = character(0), difference = numeric(0), year_delta = numeric(0))
subfacets <- unique(change_dat$subfacet)
for(i in 1:length(subfacets)){
  matches <- which(change_dat$subfacet == subfacets[i])
  years <- unique(change_dat[matches,]$year)
  if(length(years) == 1){
    next
  }
  
  ymin <- min(years)
  ymax <- max(years)
  min_stats <- change_dat[intersect(matches, which(change_dat$year == ymin)),]
  max_stats <- change_dat[intersect(matches, which(change_dat$year == ymax)),]
  comp_stats <- merge(max_stats[,c("variable", "value", "N_harmean")], min_stats[,c("variable", "value", "N_harmean")],
                      by = "variable")
  delta <- comp_stats$value.y - comp_stats$value.x
  
  hm.delta <- comp_stats$N_harmean.y - comp_stats$N_harmean.x
  
  stat_deltas <- rbind(stat_deltas, cbind(change_dat[matches[1],2:3], subfacet = subfacets[i], variable = comp_stats[,1], difference = delta, year_delta = ymax - ymin, N_harmean_delta = hm.delta))
}

stat_deltas <- merge(stat_deltas, stats[,c("Creek", "Basin", "PVA.Extinction")], by = c("Creek", "Basin"))
colnames(stat_deltas)[4] <- c("statistic")

stat_deltas_m <- reshape2::melt(stat_deltas, id.vars = colnames(stat_deltas)[-c(7:8)])



# pretty_names_tab2 <- list("Het.Hom" = "Het/Hom",
#                           "ho" = bquote(H[o]),
#                           "pi" = bquote(pi),
#                           "ws.theta" = bquote(theta[W]),
#                           "ts.theta" = bquote(theta[T]),
#                           "theta_diff" = bquote(theta[diff]),
#                           "LDNe_0.01" = bquote(LDNe[0.01]),
#                           "ne.rand" = bquote(Ne[Colony]))

pretty_names_tab2 <- c(pretty_names_tab, PVA.Extinction = "Estimated~PVA~Extinction~Risk", N_harmean_delta = "Delta*~Estimated~Harmonic~Mean~N")
vlabeller <- as_labeller(pretty_names_tab2, default = label_parsed)



# stat_deltas_m$statistic <- factor(stat_deltas_m$statistic, levels = unique(stat_deltas_m$statistic),
#                                   labels = unlist(pretty_names_tab2[match(unique(stat_deltas_m$statistic), names(pretty_names_tab2))]))
# stat_deltas_m$variable <- as.character(stat_deltas_m$variable)
# stat_deltas_m$variable[which(stat_deltas_m$variable == "PVA.Extinction")] <- "Extinction"
# stat_deltas_m$variable[which(stat_deltas_m$variable == "N_harmean_delta")] <- "Delta* Abundance"
stat_deltas_m$rate_of_change <- stat_deltas_m$difference/stat_deltas_m$year_delta

ext_plot <- ggplot(stat_deltas_m, aes(y = value, x = rate_of_change))+
  geom_point() +
  facet_grid(variable~statistic, scales = "free", labeller = vlabeller, switch = "both") + theme_bw() + 
  theme(strip.background = element_blank(), axis.title.y = element_blank(), strip.placement = "outside", 
        axis.text.x = element_text(size = 10, angle = 90), strip.text = element_text(size = 12)) +
  xlab(bquote(Delta*Statistic/Year))

ext_plot
openxlsx::insertPlot(wb, "Figure 5", width = 11, height = 8.5)
```


## Table S3 part 2

```{r}
stat_deltasc <- reshape2::dcast(stat_deltas, Creek + Basin + subfacet + year_delta + PVA.Extinction + N_harmean_delta ~ statistic, value.var = "difference", mean)
mod_stats <- stat_deltasc

# clean
mod_stats$Basin <- as.factor(mod_stats$Basin)
mod_stats$Creek <- as.factor(mod_stats$Creek)
mod_stats$N_harmean_delta <- scale(mod_stats$N_harmean_delta)
mod_stats$PVA.Extinction <- scale(mod_stats$PVA.Extinction)

run_stats <- c("pi", "ho", "ts.theta", "ws.theta", "theta_diff")

models <- data.frame(stat = numeric(length(run_stats)),
                     Ext = numeric(length(run_stats)),
                     HM = numeric(length(run_stats)),
                     Ext.model = "lm",
                     n = numeric(length(run_stats)),
                     stringsAsFactors = F)

for(i in 1:length(run_stats)){
  # prep
  tdat <- mod_stats[,c("Creek", "Basin", "N_harmean_delta", "PVA.Extinction", run_stats[i])]
  colnames(tdat)[5] <- "stat"
  bads <- which(is.na(tdat$stat) | is.infinite(tdat$stat))
  if(length(bads) > 0){
    tdat <- tdat[-bads,]
  }
  tdat$stat <- scale(tdat$stat)
  
  # run models, checking if the lmer worked for ext and running a basic lm if not
  HM <- lm(N_harmean_delta ~ stat, tdat)
  Ext <- lm(PVA.Extinction ~ stat, tdat)
  out <- c(summary(Ext)$coefficients[2,4], summary(HM)$coefficients[2,4])
  
  models[i,1] <- paste0("Delta_", run_stats[i])
  models[i,2:3] <- out
  models[i,]$n <- nrow(tdat)
}
models$holm_Ext <- p.adjust(models$Ext, "holm")
models$holm_HM <- p.adjust(models$HM, "holm")


# combine
colnames(models) <- c("Statistic", "Extinction", "Harmonic_Mean_N", "Model Type", "n", "Adjusted Extinction", "Adjusted Harmonic Mean N")
colnames(Table.S3) <- c("Statistic", "Extinction", "Harmonic_Mean_N", "Model Type", "n",
                        "Adjusted Extinction", "Adjusted Harmonic Mean N")
Table.S3 <- rbind(Table.S3, models)

# save
openxlsx::writeData(wb, "Table S3", x = Table.S3, keepNA = T)

```

## Thetas
```{r}
theta_obs <- stats[,c("Creek", "Basin", "year",
                         "ts.theta", "ws.theta", "theta_diff", "PVA.Extinction", "N_harmean")]

theta_obs$samp <- paste0(theta_obs$Basin, "_", theta_obs$Creek, "_", theta_obs$year)


## merge and plot
p2 <- ggplot(theta_obs, 
             aes(x = ts.theta, y = ws.theta, size = theta_diff, label = samp, color = N_harmean)) +
  geom_point(alpha = 0.5) + geom_abline(slope = 1, intercept = 0) + theme_bw() +
  theme(strip.background = element_blank()) + ylab(expression(theta[W])) + xlab(expression(theta[T])) +
  scale_color_viridis_c() + scale_fill_viridis_c()

p2

openxlsx::insertPlot(wb, "thetas", width = 11, height = 8.5)
```


# Save
```{r}
openxlsx::saveWorkbook(wb, "Tables_and_figures_corrected.xlsx", overwrite = T)

```

